Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2023-12-21 sections/combinatorial_semantics.tex
Soundness and completeness for rewriting

\Aleksei{Actually, in~\cite{Frobenius} a hypergraph is defined as an object of the corresponding category and labelled hypergraphs are defined as a slice category. This is needed to show adhesiveness. As our category is not adhesive, I think we can follow a more straightforward definition of a tuple.}

@@

\update{I will build below $\Ecospans$ (E for extended) instead of $ECsp(\mathcal{C})$ for arbitrary $\mathcal{C}$ with colimits to avoid talking about pushouts in general below, because we use set difference.}
\begin{definition}[Category $\Ecospans$]
    We will build the category of \textit{extended} cospans of e-hypergraphs as follows.
    Let the objects of $\Ecospans$ be only discrete e-hypergraphs (sets of edges and nodes are empty as well as both child and conflict relations) of $\catname{EHyp_{\Sigma}}$ and morphisms from $W \to Z$ be the cospans

@@

\textbf{Update: 4 Oct}

\begin{theorem}[Theorem 35~\cite{Frobenius2}]
\label{theorem:rewriting_no_frob}
    Let $R$ be any rewriting system on $\textsf{PROP}(\Sigma)$. Then,
    \[
    g \Rightarrow_{\mathcal{R}} h \;\text{ iff }\; \llangle ^{\ulcorner}g^{\urcorner} \rrangle \Rrightarrow_{\llangle \mathcal{R} \rrangle} \llangle ^{\ulcorner}h^{\urcorner} \rrangle 
    \]

    where $\llangle \cdot \rrangle : \textsf{PROP}(\Sigma) + Frob \to \catname{Csp_{D}(Hyp_{\Sigma})}$ is an isomorphism of props.
    \update{
        Last time I confused functors $\llangle \cdot \rrangle$ and $\llbracket \cdot \rrbracket$.
        Note that the result above is a weaker version of the same result for rewriting in $\textsf{PROP}(\Sigma) + Frob$.
        $Frob$ is an SMT with the signature $\Sigma_{Frob}$ which constitutes of \textit{only} Frobenius generators and terms are quotiented by both SMC laws and Frobenius equations.
        So, essentially, $Frob = \textsf{PROP}(\Sigma_{Frob}, \mathcal{E}_{Frob})$. 
        \question{Is it true that by quotienting $\textsf{PROP}(\Sigma)$ by SMC laws and quotienting $Frob$ by SMC and Frob laws we automatically quotient $\textsf{PROP}(\Sigma) + Frob$ by SMC and Frob laws?} 
    }
\end{theorem}

\begin{definition}[\;$\ulcorner \cdot \urcorner$]

    This is a syntactic operation which is defined for terms in $\textsf{PROP}(\Sigma) + Frob$ and is given below
    \[
        \tikzfig{combinatorial_semantics/bending_wires}
    \]

\begin{remark}
    We need the operation above because we defined the rewriting for cospans of hypergraphs only for cospans of the form $0 \to G \xleftarrow{} n + m$ (for $\catname{Csp_{D}(Hyp_{\Sigma})},\; \MdaCospans,\; \MdaEcospans$), but the functors take arbitrary terms (string diagrams) as input.
    This operation turns an arbitrary term into a term with no inputs.
    Now the left-hand and right-hand sides are equivalent in the sense that $f \Rightarrow_{\langle l, r \rangle} g$ if and only if $\ulcorner f \urcorner \Rrightarrow_{\langle \ulcorner l \urcorner, \ulcorner r \urcorner \rangle} 
 \ulcorner g \urcorner$ and the latter string diagram can be interpreted as $0 \xrightarrow{} G \xleftarrow{} n + m$ for which we defined rewriting above.

% \begin{proposition}
%     $m \xrightarrow{} \mathcal{G} \xleftarrow{} n$ is mda-cospan in $\MdaCospans$ iff $0 \xrightarrow{} \mathcal{G} \xleftarrow{} n + m$ is an mda-cospan.
% \end{proposition}

% In particular, the above proposition means that by fullness of $\llbracket \cdot \rrbracket$ there exists a corresponding SMC morphism for some $0 \xrightarrow{} \mathcal{G}\xleftarrow{} n + m$.
Theorem above essentially makes two routes below. In forward direction:

\[
\textsf{PROP}(\Sigma) \xrightarrow{i_1} \textsf{PROP}(\Sigma) + Frob \xrightarrow{\ulcorner \cdot \urcorner} \textsf{PROP}(\Sigma) + Frob \xrightarrow{\llangle \cdot \rrangle} \catname{Csp_{D}(Hyp_{\Sigma})}
\]

The proof of the theorem above in the forward direction then relies on the proof for rewriting in $\textsf{PROP}(\Sigma) + Frob$ and functoriality of $\llangle \cdot \rrangle$ and checking that a match is convex.

In backwards direction

\[
  \catname{Csp_{D}(Hyp_{\Sigma})} \xrightarrow{\llbracket \cdot \rrbracket}^{-1} \textsf{PROP}(\Sigma)  
\]

relies on decomposition of a hypergraph given a convex match and on fullness of

\[
    \llbracket \cdot \rrbracket : \textsf{PROP}(\Sigma) \to \MdaCospans
\]

\question{What options we have?
\begin{enumerate}
    \item Define $PROP(\Sigma)^{+}$ and $PROP(\Sigma)^{+} + Frob$
    \item Say that $n \xrightarrow{} \mathcal{G} \xleftarrow{} m$ rewrites into $n \xrightarrow{} \mathcal{F} \xleftarrow{} m$ iff $0 \xrightarrow{} \mathcal{G} \xrightarrow{} n + m$ rewrites into $0 \xrightarrow{} \mathcal{F} \xrightarrow{} n + m$.
          This gives the following route
          \[
            \textsf{PROP}(\Sigma)^{+} \xrightarrow{\llbracket \cdot \rrbracket} \MdaEcospans \xrightarrow{\ulcorner \cdot \urcorner} \MdaEcospans
          \]

          and the same in the other direction with flipped arrows.
\end{enumerate}
}
\Aleksei{TODO: discuss this}

If we quotient $\MdaCospans$ by $\llangle \mathcal{R} \rrangle$ by the theorem~\ref{theorem:rewriting_no_frob} and faithfulness and fullness of $\llbracket \cdot \rrbracket$ we get the following result.
$f = g$ in $\textsf{PROP}(\Sigma,\mathcal{E})$ iff $\llangle ^{\ulcorner} f ^{\urcorner} \rrangle = \llangle ^{\ulcorner} f ^{\urcorner} \rrangle$ in $\MdaCospans$.

\question{Do we want the same result for $\textsf{PROP}(\Sigma, \mathcal{E})^{+}$ and $\WellTypedMdaEcospans$?}

\textbf{End of update}

@@

\begin{definition}[Rewriting in $\WellTypedMdaEcospans$]
    Given $G \xleftarrow{} g_i + g_o \xleftarrow{} n_{ext} + m_{ext}$ 
    and $H \xleftarrow{} h_i + h_o \xleftarrow{} n_{ext} + m_{ext}$ in $\Ecospans$ with the outermost interfaces $n_{ext} + m_{ext}$, $G$ rewrites convexly into $H$ with the outermost interface $n_{ext} + m_{ext}$ --- notation $( G \xleftarrow{} n_{ext} + m_{ext} ) \Rrightarrow_{\mathcal{R}} (E \xleftarrow{} n_{ext} + m_{ext} )$ --- if there exists a rule $l_i + l_o \xrightarrow{} L \xleftarrow{} x_{ext} + y_{ext} \xrightarrow{} R \xleftarrow{} r_i + r_o$ with the outermost interfaces $x_{ext} + y_{ext}$ in $\mathcal{R}$ and object $C$ and cospan arrows $x_{ext} + y_{ext} \xrightarrow{} C \xleftarrow{} n_{ext} + m_{ext}$ in $\Ecospans$ such that the diagram below commutes and its marked squares are pushouts

@@

    \begin{enumerate}
        \item $m : L \to G$ is a convex down-closed match;
        \item $x_{ext} + y_{ext} \to C \to G$ is a boundary complement in the leftmost pushout.
    \end{enumerate}
\end{definition}

>>>>

% \Aleksei{Actually, in~\cite{Frobenius} a hypergraph is defined as an object of the corresponding category and labelled hypergraphs are defined as a slice category. This is needed to show adhesiveness. As our category is not adhesive, I think we can follow a more straightforward definition of a tuple.}

@@

% \update{I will build below $\Ecospans$ (E for extended) instead of $ECsp(\mathcal{C})$ for arbitrary $\mathcal{C}$ with colimits to avoid talking about pushouts in general below, because we use set difference.}

\begin{definition}[Category $\Ecospans$]
    We will build the category of \textit{extended} cospans of e-hypergraphs as follows.
    Let the objects of $\Ecospans$ be only discrete e-hypergraphs (sets of edges and nodes are empty as well as both child and conflict relations) of $\catname{EHyp_{\Sigma}}$ and morphisms from $W \to Z$ be the cospans

@@

% \textbf{Update: 4 Oct}

% \begin{theorem}[Theorem 35~\cite{Frobenius2}]
% \label{theorem:rewriting_no_frob}
%     Let $R$ be any rewriting system on $\textsf{PROP}(\Sigma)$. Then,
%     \[
%     g \Rightarrow_{\mathcal{R}} h \;\text{ iff }\; \llangle ^{\ulcorner}g^{\urcorner} \rrangle \Rrightarrow_{\llangle \mathcal{R} \rrangle} \llangle ^{\ulcorner}h^{\urcorner} \rrangle 
%     \]

%     where $\llangle \cdot \rrangle : \textsf{PROP}(\Sigma) + Frob \to \catname{Csp_{D}(Hyp_{\Sigma})}$ is an isomorphism of props.
%     \update{
%         Last time I confused functors $\llangle \cdot \rrangle$ and $\llbracket \cdot \rrbracket$.
%         Note that the result above is a weaker version of the same result for rewriting in $\textsf{PROP}(\Sigma) + Frob$.
%         $Frob$ is an SMT with the signature $\Sigma_{Frob}$ which constitutes of \textit{only} Frobenius generators and terms are quotiented by both SMC laws and Frobenius equations.
%         So, essentially, $Frob = \textsf{PROP}(\Sigma_{Frob}, \mathcal{E}_{Frob})$. 
%         \question{Is it true that by quotienting $\textsf{PROP}(\Sigma)$ by SMC laws and quotienting $Frob$ by SMC and Frob laws we automatically quotient $\textsf{PROP}(\Sigma) + Frob$ by SMC and Frob laws?} 
%     }
% \end{theorem}

% \begin{definition}[\;$\ulcorner \cdot \urcorner$]

%     This is a syntactic operation which is defined for terms in $\textsf{PROP}(\Sigma) + Frob$ and is given below
%     \[
%         \tikzfig{combinatorial_semantics/bending_wires}
%     \]

% \begin{remark}
%     We need the operation above because we defined the rewriting for cospans of hypergraphs only for cospans of the form $0 \to G \xleftarrow{} n + m$ (for $\catname{Csp_{D}(Hyp_{\Sigma})},\; \MdaCospans,\; \MdaEcospans$), but the functors take arbitrary terms (string diagrams) as input.
%     This operation turns an arbitrary term into a term with no inputs.
%     Now the left-hand and right-hand sides are equivalent in the sense that $f \Rightarrow_{\langle l, r \rangle} g$ if and only if $\ulcorner f \urcorner \Rrightarrow_{\langle \ulcorner l \urcorner, \ulcorner r \urcorner \rangle} 
%  \ulcorner g \urcorner$ and the latter string diagram can be interpreted as $0 \xrightarrow{} G \xleftarrow{} n + m$ for which we defined rewriting above.

% \end{remark}

% Theorem above essentially makes two routes below. In forward direction:

% \[
% \textsf{PROP}(\Sigma) \xrightarrow{i_1} \textsf{PROP}(\Sigma) + Frob \xrightarrow{\ulcorner \cdot \urcorner} \textsf{PROP}(\Sigma) + Frob \xrightarrow{\llangle \cdot \rrangle} \catname{Csp_{D}(Hyp_{\Sigma})}
% \]

% The proof of the theorem above in the forward direction then relies on the proof for rewriting in $\textsf{PROP}(\Sigma) + Frob$ and functoriality of $\llangle \cdot \rrangle$ and checking that a match is convex.

% In backwards direction

% \[
%   \catname{Csp_{D}(Hyp_{\Sigma})} \xrightarrow{\llbracket \cdot \rrbracket}^{-1} \textsf{PROP}(\Sigma)  
% \]

% relies on decomposition of a hypergraph given a convex match and on fullness of

% \[
%     \llbracket \cdot \rrbracket : \textsf{PROP}(\Sigma) \to \MdaCospans
% \]

% \question{What options we have?
% \begin{enumerate}
%     \item Define $PROP(\Sigma)^{+}$ and $PROP(\Sigma)^{+} + Frob$
%     \item Say that $n \xrightarrow{} \mathcal{G} \xleftarrow{} m$ rewrites into $n \xrightarrow{} \mathcal{F} \xleftarrow{} m$ iff $0 \xrightarrow{} \mathcal{G} \xrightarrow{} n + m$ rewrites into $0 \xrightarrow{} \mathcal{F} \xrightarrow{} n + m$.
%           This gives the following route
%           \[
%             \textsf{PROP}(\Sigma)^{+} \xrightarrow{\llbracket \cdot \rrbracket} \MdaEcospans \xrightarrow{\ulcorner \cdot \urcorner} \MdaEcospans
%           \]

%           and the same in the other direction with flipped arrows.
% \end{enumerate}
% }
% \Aleksei{TODO: discuss this}

% If we quotient $\MdaCospans$ by $\llangle \mathcal{R} \rrangle$ by the theorem~\ref{theorem:rewriting_no_frob} and faithfulness and fullness of $\llbracket \cdot \rrbracket$ we get the following result.
% $f = g$ in $\textsf{PROP}(\Sigma,\mathcal{E})$ iff $\llangle ^{\ulcorner} f ^{\urcorner} \rrangle = \llangle ^{\ulcorner} f ^{\urcorner} \rrangle$ in $\MdaCospans$.

% \question{Do we want the same result for $\textsf{PROP}(\Sigma, \mathcal{E})^{+}$ and $\WellTypedMdaEcospans$?}

% \textbf{End of update}

@@

\begin{definition}[Rewriting in $\WellTypedMdaEcospans$]
    Given $G \xleftarrow{} g_i + g_o \xleftarrow{} n_{ext} + m_{ext}$ 
    and $H \xleftarrow{} h_i + h_o \xleftarrow{} n_{ext} + m_{ext}$ in $\Ecospans$ with the outermost interfaces $n_{ext} + m_{ext}$, $G$ rewrites convexly into $H$ with the outermost interface $n_{ext} + m_{ext}$ --- notation $( G \xleftarrow{} n_{ext} + m_{ext} ) \Rrightarrow_{\mathcal{R}}^{+} (E \xleftarrow{} n_{ext} + m_{ext} )$ --- if there exists a rule $L \xleftarrow{} l_i + l_o \xleftarrow{} x_{ext} + y_{ext} \xrightarrow{} r_i + r_o \xrightarrow{} R$ with the outermost interfaces $x_{ext} + y_{ext}$ in $\mathcal{R}$ and object $C$ and cospan arrows $x_{ext} + y_{ext} \xrightarrow{} C \xleftarrow{} n_{ext} + m_{ext}$ in $\Ecospans$ such that the diagram below commutes and its marked squares are pushouts

@@

    \begin{enumerate}
        \label{dpoi:assumptions}
        \item $m : L \to G$ is a convex down-closed match;
        \item $[m(v_i)) = [m(v_j))$ for all $v_i,v_j$ that have a pre-image in $x_{ext} + y_{ext}$
        \item $\consistency^{\mu}(m(v_i)) = \consistency^{\mu}(m(v_j))$ for all $v_i,v_j$ that have a pre-image in $x_{ext} + y_{ext}$
        \item $x_{ext} + y_{ext} \to C \to G$ is a boundary complement in the leftmost pushout.
    \end{enumerate}
\end{definition}

\begin{remark}
The conditions 2 and 3 are needed to guarantee the existence of the pushout complement.
\end{remark}
